Theories of deliberated judgment

### 2022-08-06 simple.tex
Brought old defs

\subsection{General theories of deliberated choice}
\label{sec:dynamic}
A theory of deliberated choice presents a claim $C$ under the form of a constraint on ${\ileadsto}$.
That claim

@@

\subsection{Definition}
\label{sec:descrdef}
In this section, a normative theory of \ac{DJ} $(\allprops, \allargs, I, {\allleadsto})$ is supposed given.
A descriptive theory $\gamma$ for a normative theory of \ac{DJ} is defined as a tuple $(\gC, \gPhi)$, where $\gC \subseteq \powerset{I × \allhist × \Phi}$ is a falsifiable claim, $\gPhi \subseteq \Phi$ are the propositions that the theory claims are in the \acp{DJ} of the individuals, and such that its falsifiable claim implies its convincingness.
\commentYM{dans la mesure où, à ce stade, je ne comprends pas bien ce que contient une "normative theory", la séparation avec l' "empirical theory" n'est pas claire pour moi} 
These concepts are now defined formally.
\commentYM{puisque la théorie empirique est une théorie empirique D'UNE CERTAINE THEORIE NORMATIVE, je m'attendrais  à ce qu'on voit apparaître la théorie normative dans la formule de la théorie empirique, non ?}

@@

\begin{example}[A theory for $P$]
	\label{ex:twolottstrings}
	Define a normative theory as in \cref{ex:twolott}, with an extended set of arguments: define $\allargs$ as the set of all strings. (This correspondingly changes $\allhist$ and thus $\ileadsto$, which must be extended to its new domain.) 
	Define $\ar_0$ as in \cref{ex:twolott}.
	Define $\ar_\text{for$P$}$ as the text “We recommend to adopt $P$: it has a much higher probability of winning.”

>>>>

\subsection{Argumentative theories of deliberated choice}
\label{sec:dynamic}
An argumentative theory of \ac{DC} is defined as a tuple $(\gleadsto, \dg)$, where ${\gleadsto} \subseteq \allargs × \allprops$ is an initial argumentation relation and $\dg$ is a defense relation; satisfying bounded depth.

\subsection{Old definition}
\label{sec:descrdef}
In this section, a normative theory of \ac{DJ} $(\allprops, \allargs, I, {\allleadsto})$ is supposed given.
These concepts are now defined formally.
\commentYM{puisque la théorie empirique est une théorie empirique D'UNE CERTAINE THEORIE NORMATIVE, je m'attendrais  à ce qu'on voit apparaître la théorie normative dans la formule de la théorie empirique, non ?}

An initial argumentation relation ${\gleadsto} \subseteq \allargs × \allprops$ defines the set of supported  propositions and, for each supported propositions, the set of initial arguments for these propositions. 
More precisely, let $\gpropse$, the range of ${\gleadsto}$, be called the propositions \emph{supported} by the descriptive theory.
Given such a supported proposition $\prop \in \gpropse$, let $\gleadstoinv(\prop) = \set{\ar \in \allargs \suchthat \ar \gleadsto \prop}$ be called the initial arguments in favor of $\prop$.

A defense relation $\dg \subseteq (\allargs × \allhist) × \allargs$ is a right-unique relation, thus given any $\ar \in \allargs, \hist \in \allhist$, $\dg(\ar, \hist)$ is empty or is a singleton. %, and must satisfy $\ar \notin \histend ⇒ \dg(\ar, \hist) = \emptyset$. 
As is usual for right-unique relations, I write $\dg(\ar_0, \hist) = \ar_1$ instead of $\dg(\ar_0, \hist) = \set{\ar_1}$.
Say that $\ar_1$ defends $\ar_0$ iff $\exists \hist \suchthat \dg(\ar_0, \hist) = \ar_1$.
Define $\dg(\ar) = \dfp(\ar) = \dg(\ar, \allhist) = \set{\dg(\ar, \hist) \suchthat \hist \in \allhist} \subseteq \allargs$ as all the defenders of $\ar$.
The relation $\dg$ defines the argumentation strategy of the theory. Informally speaking, its role is to adapt the arguments the theory produces to the reactions of the individual it is trying to convince. Assuming the theory has played an argument $\ar_0$ and a path $\hist$ challenges some of the conclusions this argument supports (as judged by the individual), $\dg(\ar_0, \hist)$ is the argument that the theory plays as a counter-argument to the path $\hist$.
%The condition that when $\dg(\ar_0, \hist)$, $\ar_0 \in \histend$ is imposed for simplicity: as will appear, an argument $\ar_0$ needs defense only when it is challenged by some path $\hist \suchthat \ar_0 \in \histend$, thus, letting $\dg$ define counter-arguments to paths not satisfying this condition would be of no use.

This mechanism permits to define defenders of defenders of an argument, $\dg(\dg(\ar))$, defenders of defenders of defenders of an argument, … 
For the results to be presented here to hold, an empirical theory must be a priori restricted in the maximal “depth” of its argumentation strategy, in the following sense. 
\commentYM{ok soit, ça semble raisonnable, mais tu n'as pas de raison plus forte, j'adopterais une formulation plus modeste et ouverte, du genre "ça semble souhaitable"}
Given any set $S$ and function $f: S → S$, let $f^0$ represent the identity function, and given $k \in \N$, let $f^{k + 1} = f \circ f^k$.
The bounded depth condition on a descriptive theory imposes that
$\exists k \in \N \suchthat (\dg)^k(\gleadstoinv(\allprops)) = \emptyset$.
The smallest such $k$ is called the depth of the theory.

Here is an example of a theory that does not use an argumentation strategy, but simply plays an argument that it claims is decisive for everybody.
\begin{example}[Constant argumentation]
	Given the normative theory of \cref{ex:twolott}, define ${\gleadsto} = \set{(\ar_0, D > P)}$ and $\dg = \emptyset$. This defines a descriptive theory of depth $1$ for that normative theory.
\end{example}

\begin{example}[A theory for $P$]
	\label{ex:twolottstrings}
	Define a normative theory as in \cref{ex:twolott}, with an extended set of arguments: define $\allargs$ as the set of all strings. (This correspondingly changes $\allhist$ and thus $\ileadsto$, which must be extended to its new domain.) Define $\ar_0$ as in \cref{ex:twolott}.
	Define $\ar_\text{for$P$}$ as the text “We recommend to adopt $P$: it has a much higher probability of winning.” 
	Define ${\gleadsto} = \set{(\ar_\text{for$P$}, P > D)}$.
	In order to define a counter-argument to $\ar_0$, given a real number $r \in \R$, define $\ar^r_\text{non-linear}$ as the text “$D$ is better than $P$ in terms of expected gain, but it is in general absurd to reason in terms of expected gain. For example, consider a choice between winning $r$ dollars with probability $0.8$ and winning $2r$ dollars with probability $0.4$. They have the same expected gain, but the first one is clearly preferrable. Therefore, we recommend to rather pick $P$ as it presents a much better chance of earning money”. (The value of $r$ and $2r$ is to be used in the relevant places in  the text.)
	Define $\dg = \set{(\ar_\text{for$P$}, \hist, \ar^{10^6}_\text{non-linear}) \suchthat \hist \in \allhist \land \ar_0 \in \hist}$. This strategy anticipates a possible attack using $\ar_0$ against $\ar_\text{for$P$}$, and prepares $\ar^{10^6}_\text{non-linear}$ as a counter-argument.
	This defines a descriptive theory of depth $2$.
\end{example}

The following less trivial example of a descriptive theory illustrates a theory that claims to be able to answer an infinite number of arguments. This also shows that the bounded depth constraint does not imply finiteness of the relation $\dg$.
\begin{example}[Infinite $\dg$]
	\label{ex:infinite}
	Define a normative theory and the arguments $\ar_0$ and $\ar^r_\text{non-linear}$ for any $r \in \R$ as in \cref{ex:twolottstrings}.
	Define ${\gleadsto} = \set{(\ar_0, D > P)}$.
	Define $\ar_\text{linear}$ as the text “We agree that when the money amounts involved are large and the game is not repeated, there is no good reason to consider only expected gains when choosing a lottery. But in our case, the money amounts are small and the game is repeated. Therefore, expected gain is the right indicator to consider for a systematic and sound decision strategy.”
	Define $+$ as the concatenation operator over strings.
	Define $\dg = \set{(\ar_0, \ar^r_\text{non-linear}, \ar_0 + \ar_\text{linear}) \suchthat r \in \R}$. This strategy sets up the argument $\ar_0 + \ar_\text{linear}$ as an answer to any argument of the form described in \cref{ex:twolottstrings}, arguing for non-linearity of utility of money.
	This defines a descriptive theory of depth $2$.
\end{example}

\subsection{Validity and falsification}
\label{sec:validity}
An empirical theory claims to be valid, in the following sense.
\begin{definition}[Validity]
	An empirical theory is \emph{valid} iff $\forall i \in I, \prop \in \gpropse: \prop \in \iprops$.
\end{definition}
To achieve this goal, it claims that it is able to produce decisive arguments for its supported propositions.
These arguments can vary according to the individual $i \in I$.
Given an individual, the theory claims that a decisive argument will be systematically found by starting from an initial argument given by $\gleadsto$, and following a chain of defense defined by $\dg$ as required, depending on the paths that challenge the arguments produced along the way, as judged by the individual $i$. The property that a theory manages to systematically produce decisive arguments in such a way is named Convincingness. Defining it precisely requires to define first the notion of required arguments.

Given $i \in I, \prop \in \gpropse$, define $\dgip(\ar) = \set{\dg(\ar, \hist) \suchthat \ar \notin \histend \land \hist \nileadsto \phi} \subseteq \dg(\ar)$, representing the subset of defenders of $\ar$ required for convincing $i$ of $\prop$.
Given any set $S$ and function $f: S → S$, let $f^0$ represent the identity function, and given $k \in \N$, let $f^{k + 1} = f \circ f^k$.
Define $\gargs \subseteq \allargs$ as the transitive closure of ${\gleadstoinv}(\prop)$ under $\dgip$: $\gargs = \cup_{k \in \N} (\dgip)^k({\gleadstoinv}(\prop))$. The set $\gargs$ is the set of arguments required to convince $i$ of $\phi$.
\begin{example}[Required arguments]
	\label{ex:required}
	Continuing \cref{ex:infinite}, define the subset of individuals that find some argument $\ar^r_\text{non-linear}$ convincing in presence of $\ar_0$ as the set of $i \in I$ such that $\exists \hist \in \allhist \suchthat \ar_0 \in \histend \land \ar^r_\text{non-linear} \in \range(\hist) \land \hist \nileadsto D > P$.
	We see that for any $i$ in that set of individuals, $\gargs[D > P] = \set{\ar_0, \ar_0 + \ar_\text{linear}}$, and for any $i$ in the complementary set of individuals, $\gargs[D > P] = \set{\ar_0}$.
\end{example}

Given $\ar \in \gargs$, say that $\ar$ $i$-defends $\phi$ iff $\forall \hist \in \allhist \suchthat \ar \in \histend \land \hist \nileadsto \phi: d(\ar, \hist) ≠ \emptyset$.
\begin{definition}[Convincingness]
	An empirical theory is \emph{convincing} iff $\forall i \in I, \prop \in \gpropse, \ar \in \gargs: \ar$ $i$-defends $\prop$.
\end{definition}
\begin{example}[Convincingness]
	Pursuing with \cref{ex:required}, that theory will be convincing iff either $i$ finds some argument $\ar^r_\text{non-linear}$ convincing in presence of $\ar_0$ and both $\ar_0$ and $\ar_0 + \ar_\text{linear}$ $i$-defend $D > P$, or $i$ does not find any argument $\ar^r_\text{non-linear}$ convincing in presence of $\ar_0$ and $\ar_0$ $i$-defends $D > P$.

	$\ar_0$ $i$-defends $D > P$ iff any argumentative path $\hist$ that contains $\ar_0$ towards its end and that does not lead $i$ to accept $D > P$ is such that $\dg(\ar_0, \hist) ≠ \emptyset$, or equivalently, is such that the theory has some argument to answer to $\hist$, or still equivalently, is such that it contains some $\ar^r_\text{non-linear}$.
	This property mandates that no argumentative path has been “forgotten” in $\dg$: if some path is found that attacks $\ar_0$ successfully (thus $\ar_0 \in \histend \land \hist \nileadsto D > P$), and if the theory has nothing to answer to it (thus $\dg(\ar_0, \hist) = \emptyset$), then the theory is not Convincing.
\end{example}

It is now necessary to verify that the claim of convincingness is, first, falsifiable, and second, sufficient to yield validity.

Let us first clarify what falsifiability formally means in our context. So far, convincingness is defined only for the “real” relation $\leadsto$; we now need it defined more generally. Given any relation $R \subseteq I × \allhist × \Phi$, say that a descriptive theory is convincing under $R$ whenever it is convincing, as defined above, but replacing $\leadsto$ by $R$. 
We can now speak counterfactually of the theory being convincing, or not, assuming the observable relation is $R$ instead of $\leadsto$. 
Observations of the relation $\leadsto$ can in general include positive statements, of the form $\hist \ileadsto \phi$, and negative ones, of the form $\hist \nileadsto \phi$. (One way to obtain such negative observations is to use an incompatible proposition $\phi'$ and deduce from $\hist \ileadsto \phi'$ that $\hist \nileadsto \phi$.) A relation $R \subseteq I × \allhist × \Phi$ is compatible with the positive observations $O \subseteq I × \allhist × \Phi$ and the negative observations $O' \subseteq I × \allhist × \Phi$ iff $O \subseteq R \land R \cap O' = \emptyset$.
The claim of convincingness is said to be falsifiable iff there exists finite set of observations $O, O' \subseteq I × \allhist × \Phi$ such that $\forall R \subseteq I × \allhist × \Phi \suchthat O \supseteq R \land R \cap O' = \emptyset$, the theory is not convincing under $R$. 
In such a case, no possible relation $R$ compatible with the observations $O, O'$ make the theory convincing; in other words, the observations $O, O'$ prove that the theory is not convincing.
%\begin{example}[Two claims]
%	Given a normative theory and $\phi \in \Phi$, the claim $C_1 = \set{{\leadsto} \subseteq I × \allhist × \Phi \suchthat \forall i \in I, \exists \ar \in \allargs \suchthat \ar \ileadstost \phi}$ corresponds to claiming that for all individuals $i \in I$, some argument is $i$-decisive for $\phi$. If the set of arguments is infinite, $C_1$ is not falsifiable.  
%	Given $\ar \in \allargs$ and $\phi \in \Phi$, the set $C_2 = \set{{\leadsto} \subseteq I × \allhist × \Phi \suchthat \forall i \in I, \ar \ileadstost \phi}$ denotes the claim that all individuals will consider the argument $\ar$ as decisive for $\phi$. This claim is falsified by any observation such that $\hist \nileadsto \phi$ for some $\hist \in \allhist$ with $\ar \in \histend$, thus, $C_2$ is falsifiable.
%\end{example}

The following theorem states that the descriptive theories defined here are falsifiable, unless they state nothing about anybody’s \ac{DJ}. \Cref{sec:prooffals} proves it and describes the set of falsifiers of such theories.
This theorem implicitly assumes full observability, that is, it supposes that any positive or negative observations are in principle observable and thus available for falsification attempts. Further works should investigate to which extent this hypothesis can be relaxed.
\begin{theorem}
	\label{th:convfals}
	The claim of convincingness is falsifiable iff ${\gleadsto} ≠ \emptyset$.
\end{theorem}
It can also be seen that if a theory is not constrained in its depth, its claim of convincingness may be unfalsifiable, thus the hypothesis of finite depth is required to obtain this theorem. The formal proof of this fact is omitted because it is not especially enlightening; that this fact holds can be seen by examining the use of this hypothesis in the proof of \cref{th:convfals}.

Only remains now to prove that convincingness, together with normative adequacy, is sufficient to obtain validity. The proof is also provided in \cref{sec:prooffals}.
\begin{theorem}
	\label{th:convvalid}
	A normatively adequate empirical theory that is convincing is valid.
\end{theorem}

\subsection{Effectively comparing models}
\label{sec:compare}
Recall from \cref{sec:coh} that $\incompat = \set{(\prop, \prop') \suchthat \forall \hist \in \allhist: \hist \ileadsto \prop ⇒ \hist \nileadsto \prop'}$ denotes the set of pairs of incompatible propositions.
\commentOC{Does this mandate completeness?}
Given two empirical theories $\gamma$ and $\delta$ for a given normative theory, say that they disagree about $(\prop, \propbar) \in \incompat$ iff $\prop \in \gpropse[\gamma] \land \propbar \in \gpropse[\delta]$. Say that the theories are incompatible iff there is at least one pair $(\prop, \propbar) \in \incompat$ that they disagree about.

A path $\histp$ starts with a path $\hist$ iff $\card{\histp} ≥ \card{\hist}$ and $\forall k \in \intvl{1, \card{\hist}}: \histp_k = \hist_k$.
An increasing sequence of paths is a finite sequence $({\hist}^k)_{k \in \intvl{1, N}}$, for some $N \in \N$, of paths, 
such that $\forall k \in \intvl{2, N}, {\hist}^k$ starts with ${\hist}^{k-1}$.

\begin{theorem}
	\label{th:proc}
	Given two incompatible empirical theories for a given normative theory, $\gamma = (\gleadsto[\gamma], \gbeats[\gamma], \alldargs_\gamma, \dbeats[\gamma], +_\gamma)$ and $\delta = (\gleadsto[\delta], \gbeats[\delta], \alldargs_\delta, \dbeats[\delta], +_\delta)$ and given any $i \in I$, there exists a procedure that falsifies one of these theories, queries only $i$, and queries $i$ only about some increasing sequence of paths.
\end{theorem}
That procedure is described in \cref{sec:proc}.

@@

\subsection{Old definition appendix}
\label{sec:descrdefapdx}
In this section, a normative theory of \ac{DJ} $(\allprops, \allargs, I, {\allleadsto})$ is supposed given.
A descriptive theory $\gamma$ for a normative theory of \ac{DJ} is defined as a tuple $(\gC, \gPhi)$, where $\gC \subseteq \powerset{I × \allhist × \Phi}$ is a falsifiable claim, $\gPhi \subseteq \Phi$ are the propositions that the theory claims are in the \acp{DJ} of the individuals, and such that its falsifiable claim implies its convincingness.
\commentYM{dans la mesure où, à ce stade, je ne comprends pas bien ce que contient une "normative theory", la séparation avec l' "empirical theory" n'est pas claire pour moi} 
These concepts are now defined formally.
\commentYM{puisque la théorie empirique est une théorie empirique D'UNE CERTAINE THEORIE NORMATIVE, je m'attendrais  à ce qu'on voit apparaître la théorie normative dans la formule de la théorie empirique, non ?}

@@

\begin{example}[A theory for $P$]
	Define a normative theory as in \cref{ex:twolott}, with an extended set of arguments: define $\allargs$ as the set of all strings. (This correspondingly changes $\allhist$ and thus $\ileadsto$, which must be extended to its new domain.) 
	Define $\ar_0$ as in \cref{ex:twolott}.
	Define $\ar_\text{for$P$}$ as the text “We recommend to adopt $P$: it has a much higher probability of winning.”
